---
title: "N₂O Emissions in Sub-Saharan Africa"
author: "Ouma Turry"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction
This document analyzes N₂O emissions from cropland soils in Sub-Saharan Africa using field measurement data.


# Load Required Libraries
```{r}
library(ggplot2)
library(zen4R)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(leaflet)
library(plotly)
library(ggpubr)
library(readxl)
library(RCurl)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/turryouma/FAIRLEAP/Workflow")
data_fert <- readxl::read_excel("/Users/turryouma/FAIRLEAP/Workflow/data/fertemissions.xlsx")
data_fert
main_sum <- read.csv("/Users/turryouma/FAIRLEAP/Workflow/data/main_sum_delimit.csv")
main_sum
```

# Load and Prepare Data
```{r}
# Load the world map shapefile
world <- ne_countries(scale = "medium", returnclass = "sf")

# Filter for Sub-Saharan Africa
ssa <- world %>%
  filter(continent == "Africa" & !name %in% c("Morocco", "Algeria", "Tunisia", "Libya", "Egypt"))

# Read in data
main_sum$mean <- as.numeric(main_sum$mean)

# Categorize emissions
main_sum <- main_sum %>%
  mutate(
    emission_range = cut(mean, 
                         breaks = c(-Inf, 0.5, 1, 1.5, 2, 3, 4, Inf), 
                         labels = c("< 0.5", "0.5–1", "1.1–1.5", "1.6–2", 
                                    "2.1–3", "3.1-4", ">4.1"))
  )
```

# Convert Data to Spatial Format
```{r}
data_sf <- st_as_sf(main_sum, coords = c("longitude", "latitude"), crs = 4326)
```

# Visualization: N₂O Emissions in SSA
```{r}
ssa_countries <- ssa %>%
  mutate(Study_Availability = ifelse(name %in% unique(data$Country), "Yes", "None"))

# Calculate centroids for labels
ssa_centroids <- st_centroid(ssa_countries)
ssa_centroids <- cbind(ssa_centroids, st_coordinates(ssa_centroids))

# Filter fertilizer-induced emissions
fertind_sf <- filter(data_sf, fertilizer_induced == "yes")

ssa_sf <- ssa_countries %>%
  left_join(df_clean %>% select(Country, fertilizer_induced, latitude, longitude), 
            by = c("name" = "Country")) %>%
  mutate(Study_Availability = ifelse(!is.na(fertilizer_induced), "Study", "No Study"))

# Summarize emissions data by country
country_summary <- fertind_sf %>%
  group_by(Country, Location) %>%
  summarize(
    mean_emissions = mean(mean, na.rm = TRUE), # Average emissions
    min_emissions = min(min, na.rm = TRUE),   # Minimum emissions
    max_emissions = max(max, na.rm = TRUE),   # Maximum emissions
    geometry = st_union(geometry) %>% st_centroid() # Country centroid for plotting
  )
```


```{r}
#############
dist_map <- ggplot() +
  # Layer 1: All SSA countries in grey
  geom_sf(data = layer1_data, fill = "grey", color = "white") +
  
  # Layer 2: Countries where studies have been conducted (light blue)
  geom_sf(data = layer2_data, fill = "#8BC34A", color = "white") +
  
  # Layer 3: Countries with fertilizer-induced emissions (green)
  geom_sf(data = layer3_data %>% filter(fertilizer_induced == "yes"), fill = "#388E3C", color = "white") +
  
  # Layer 4: Emission points
  geom_sf(data = country_summary, aes(color = mean_emissions, size = mean_emissions)) +
  
  scale_color_gradientn(
    colors = c("yellow", "orange", "#d53e4f", "#3288bd","blue", "purple"), 
    breaks = c(0, 1, 2, 3),  # Show only these values in the legend
    limits = c(0, 3),  # Ensure the gradient range stops at 2.7
    name = expression("Mean Emissions of " ~ N[2]*O ~ "(kg N"["2"]*"O-N/ha/year)")
  ) +
  
  scale_size_continuous(
    range = c(2, 8), 
    guide = "none"  # Remove size legend
  ) +
  
  # Layer 5: Country labels
  geom_text(data = ssa_centroids, aes(X, Y, label = name), size = 3, color = "black") +
  
  # Title and caption
  labs(title = expression("Studies on" ~ N[2]*O~ "Emissions in Sub-Saharan Africa"),
       subtitle = "Layered visualization: SSA countries, studies conducted, fertilizer-induced emissions, and emission points",
       caption = expression("Data Source: Peer reviewed SSA studies on" ~ N[2]*O~ "emissions from croplands")) +
  
  # Map theme
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 14),  
    legend.text = element_text(size = 14),   
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    plot.caption = element_text(size = 12)
  ) +
  coord_sf(expand = FALSE)
dist_map
```

# Save Plot
```{r}
ggsave("dist_map.png", plot = dist_map, width = 12, height = 8, dpi = 300)
```

```{r}

# Rename columns for clarity 
colnames(data_fert) <- c("Study", "Country", "Treatment", "Rate", "Emission", "Unit")

# Convert `Rate` to a factor with a defined order (from Control to highest range)
data_fert$Rate <- factor(data_fert$Rate, levels = c("Control", "<50", "51-100", "101-150", ">150"))

# Calculate the mean emissions for each group
means <- data_fert %>%
  group_by(Rate) %>%
  summarise(mean_emission = mean(Emission, na.rm = TRUE))

# # Get a list of studies used in the analysis
# study_list <- unique(data$Study)
# study_caption <- paste("Studies included:", paste(study_list, collapse = ", "))
# Get a list of studies used in the analysis
study_list <- unique(data_fert$Study)

# Define the number of studies per line
studies_per_line <- 6

# Split the study list into chunks
study_caption_lines <- split(study_list, ceiling(seq_along(study_list) / studies_per_line))

# Combine the chunks into separate lines
study_caption <- paste("Studies included:", 
                       paste(sapply(study_caption_lines, function(x) paste(x, collapse = ", ")), 
                             collapse = "\n"))
```


```{r}
#by country
# Interaction plot
emfert <- ggplot(data_fert, aes(x = Rate, y = Emission, color = Country, group = Country)) +
  geom_boxplot(aes(fill = Country), alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.8) +
  labs(
    x = "Nitrogen fertilizer application rate (Kg/ha)",
    y = expression("Emission (Kg"  ~ N[2]*O ~ "(kg N"["2"]*"O-N/ha/year)"),
    title = expression("Interaction of Fertilizer Rate and Country on" ~ N[2]*O ~ "Emissions")
  ) +
  theme_minimal(base_size = 14)
emfert
```

```{r}
# Create the boxplot
comp_means <- ggplot(data_fert, aes(x = Rate, y = Emission, fill = Rate)) +
  geom_boxplot(outlier.shape = NA) +  # Boxplot without outliers
  geom_jitter(aes(color = Rate), width = 0.2, size = 2, alpha = 0.8) +  # Add jittered points
  scale_fill_brewer(palette = "Set2") +  # Set color palette for boxes
  scale_color_brewer(palette = "Set2") +  # Set color palette for points
  labs(
    x = "Nitrogen fertilizer application rate (Kg/Ha)",
    y = expression("Total " ~ N[2]*O ~ " emission (kg N"["2"]*"O-N/ha/year)"),
    
    caption = study_caption  # Add the studies as a caption
  ) +
  theme_minimal(base_size = 14) +  # Set minimal theme with larger font size
  theme(
    legend.position = "none",  # Remove legend
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    plot.caption = element_text(hjust = 0)  # Align caption to the left
  ) +
  # Add mean values on top of the boxplots
  geom_text(
    data = means,
    aes(x = Rate, y = mean_emission + 0.5, label = round(mean_emission, 2)),
    color = "black",
    size = 4
  ) +
  # Optional: Add pairwise statistical comparisons
  stat_compare_means(
    method = "t.test",
    comparisons = list(
      c("Control", "<50"),
      c("<50", "51-100"),
      c("51-100", "101-150"),
      c("101-150", ">150")
    ),
    label.y = c(6.5, 7, 7.5, 8),
    size = 4
  )

comp_means
```
```{r}
fert_country <-ggplot(data_fert, aes(x = Rate, y = Emission, fill = Rate)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color = Rate), width = 0.2, size = 2, alpha = 0.8) +
  labs(
    x = "Nitrogen fertilizer application rate (Kg/ha)",
    y = expression("Total " ~ N[2]*O ~ " emission (kg N"["2"]*"O-N/ha/year)"),
    title = "" ~ N[2]*O ~ " Emissions by Fertilizer Rate for Each Country"
  ) +
  theme_minimal(base_size = 14) +
  facet_wrap(~Country, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2")
fert_country
```

```{r}
#linear fit
# 1. Calculate regression statistics for each country
stats <- data_fert %>%
  group_by(Country) %>%
  do({
    fit <- lm(Emission ~ Treatment, data = .)
    summary_fit <- summary(fit)
    data.frame(
      Country = .$Country[1],
      intercept = coef(fit)[1],
      slope = coef(fit)[2],
      r_squared = summary_fit$r.squared,
      p_value = coef(summary_fit)[2, 4]
    )
  })

# 2. Merge regression statistics back into the dataset for annotation
data_fert2 <- data_fert %>% filter(Country != "Madagascar")
data_with_stats <- data_fert2 %>%
  left_join(stats, by = "Country")
# Fit linear models for each country and extract the required statistics
country_stats <- data_with_stats %>%
  group_by(Country) %>%
  do({
    fit <- lm(Emission ~ Treatment, data = .)
    tibble(
      Country = unique(.$Country),
      r_squared = summary(fit)$r.squared,  # Use R² instead of adjusted R²
      intercept = fit$coef[[1]],
      p_value = summary(fit)$coef[2, 4],
      x_pos = min(.$Treatment),  # Position label based on min Treatment (for left side)
      y_pos = max(.$Emission)   # Position label based on max Emission (for top side)
    )
  })
```

```{r}
scatter_emiss <- ggplot(data_with_stats, aes(x = Treatment, y = Emission)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, fill="#69b3a2",color = "darkgreen") +
  facet_wrap(~ Country, scales = "free") +  # Facet by country with free y-axis scales
  geom_label(data = country_stats, 
             aes(x = x_pos, y = y_pos, 
                 label = paste("R² = ", signif(r_squared, 5),
                               
                               "\nP =", signif(p_value, 5))),
             hjust = 0, vjust = 1, size = 3, alpha = 0.3)  # Position the label at top-left

scatter_emiss
```





# Conclusion
This analysis visualizes N₂O emissions data from Sub-Saharan Africa, highlighting areas where studies have been conducted and variations in emissions intensity.
